Inference using context-dependent models of mutation via bounds on information propagation

### 2021-12-07 writeups/context-dependence.tex
notes from chat

\begin{figure}
    \begin{center}
        \includegraphics{overview}
    \end{center}
    \caption{
        An overview of the method for a neighbor-dependent model.
        The red triangle represents the ``range of influence'',
        i.e., possible locations at which a mutations is likely
        to affect the mutation outcome at
        a site of interest (the site at the bottom corner of the triangle).
        The blue T shape is called a \emph{T-mer},
        which is how our approximation to the process is parameterized.
        The central segment of the T is called the \emph{base} of the T-mer,
        and the key building block computes the transition probability
        for this base
        conditioned on the ancestral state at the \emph{overhang}
        of the top of the T.
        \label{fig:overview}
    }
\end{figure}

@@

We can then simply express our likelihood as
\[
    \like(y \mid x) = \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
\]
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}
See Appendix~\ref{ss:approx_pf} for a more rigorous statement and proof,
and the next section for some intuition.

@@

Using the R package expm \citep{R_expm}, this makes computation quite feasible:
with four possible states and $\ell=2$ and $m=1$, so that $G(n)$ is a $1024 \times 1024$ matrix,
computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
%EM In the above, don't we only provide the recipe for the full likelihood when m = 2 ell + 1?
Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
(and much longer for the whole matrix $e^{t G(8)}$).

@@

Posterior medians were within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
As seen in Table~\ref{tab:ising_coverage},
%EM I think this credible interval statement is interesting. So it's definitely this narrowness vs just having the estimate be off?
credible intervals were somewhat too narrow for shorter T-mers,
but $(6,2,2)$ T-mers were long enough with sufficient overhang to produce well-calibrated posteriors.

>>>>

\begin{figure}
    \begin{center}
        %EM Make width like in figure 2. 2 and 2.
        %EM Bottom of triangle should be three bases.
        \includegraphics{overview}
    \end{center}
    \caption{
        An overview of the method for a neighbor-dependent model.
        The red triangle represents the ``range of influence'',
        i.e., possible locations at which a mutations is likely
        to affect the mutation outcome at
        a site of interest (the site at the bottom corner of the triangle).
        The blue T shape is called a \emph{T-mer},
        which is how our approximation to the process is parameterized.
        The central segment of the T is called the \emph{base} of the T-mer,
        and the key building block computes the transition probability
        for this base
        conditioned on the ancestral state at the \emph{overhang}
        of the top of the T.
        \label{fig:overview}
    }
\end{figure}

@@

We can then simply express our likelihood as
\[
    \like(y \mid x) = \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
\]
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}

%EM It seems reasonable that we should be able to get away using a smaller T-mer.
%EM Question: how much context should we include; argument in appendix.
%EM Would be nice to show convergence of full likelihood eventually.

See Appendix~\ref{ss:approx_pf} for a more rigorous statement and proof,
and the next section for some intuition.

@@

Using the R package expm \citep{R_expm}, this makes computation quite feasible:
with four possible states and $\ell=2$ and $m=1$, so that $G(n)$ is a $1024 \times 1024$ matrix,
computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively
(and much longer for the whole matrix $e^{t G(8)}$).

@@

Posterior medians were within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
As seen in Table~\ref{tab:ising_coverage},
credible intervals were somewhat too narrow for shorter T-mers,
but $(6,2,2)$ T-mers were long enough with sufficient overhang to produce well-calibrated posteriors.
